Left-linear rules in adhesive categories

### 2024-02-10 icalp/fusione.tex
very tame

\documentclass[a4paper,UKenglish,cleveref,pdftex, thm-restate,numberwithinsect,anonymous]{lipics}
\newif\ifreport
\reporttrue
%\reportfalse

% full includes some additional material
\newif\iffull
\fulltrue
\fullfalse

% additional text
\iffull
\newcommand{\full}[1]{{\color{blue}#1 full}}
%\newcommand{\short}[1]{}
\else 
\newcommand{\full}[1]{}
%\newcommand{\short}[1]{#1}
\fi

@@

\def\R{\mathsf{R}}
\def\B{\textbf {\textup{B}}}
\def\C{\textbf {\textup{C}}}
\def\D{\textbf {\textup{D}}}
\def\X{\textbf {\textup{X}}}
\def\Y{\textbf {\textup{Y}}}

@@

\title{Left-linear rules in adhesive categories}

\author{Paolo Baldan} 
{Department of Mathematics, University of Padova, Italy}
{baldan@math.unipd.it}{}{}
\author{Davide Castelnovo}
{Department of Mathematics, University of Padova, Italy}
{davide.castelnovo@math.unipd.it}{}{}
\author{Andrea Corradini}
{Department of Computer Science, University of Pisa, Italy}
{andrea.corradini@unipi.it}{}{}

@@

\begin{abstract}
  When two sequences of steps executed by a computational device can
  be considered equivalent?  This is a relevant question if one
  considers concurrent and distributed systems, since the independence
  of two steps means that they could be executed on any order, and
  potentially in parallel.
  % 
  We investigate the issue in the context of adhesive categories, a
  very general instance of the DPO approach to rewriting. More
  precisely, we focus on left-linear rules, which not only consume,
  preserve and generate entities, as linear rules do, but may
  ``merge'' parts of the state.
  % 
  An apparently minimal, yet technically hard enhancement, which
  allows to properly extend the formalisms that can be modelled in the
  adhesive framework, and for which we prove a Church-Rosser property.
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  The canonical notion of independence used for linear rewriting
  systems does not work out of the box, but it requires the
  restriction to suitable well-behaved classes of left-linear
  rewriting systems.  Additionally, a fundamental characterisation of
  the notion of independence fails when moving to left-linear
  rewriting systems. This requires a substantial revision of the
  theory as in the linear case, a number of basic properties,
  essential in the development of the theory of concurrency, were
  derived from such a characterisation.
 \end{abstract}

@@

Here we observe that, in general, both properties can fail: sequential
independent steps could be not switchable and different independent
pairs could exist leading to different switches of the same rewriting
steps.  A refined notion of independence can be considered to ensure
that independent rewriting steps can be switched, based in the concept
of a so-called filler. The problem was identified already
in~\cite{baldan2011adhesivity} where a suitable subclass of adhesive
categories is identified where independence as induced by a filler
coincide with sequential independence (as induced by an independence
pair). Here this is extended to $\mathcal{M}$-adhesive categories.

@@

\todo{P: qui probabilmente si puo' compattare tenendo solo cio' che si usa}
\todo{Si usa tutto per la classe B+ [DC]}
$\mathcal{M}$-adhesivity is well-behaved with respect to  the comma construction \cite{mac2013categories}, as shown by the following theorem.
\begin{theorem}
[\cite{ehrig2006fundamentals,lack2005adhesive}]
\label{lem:comma}
  Let $\A$ and $\B$ be respectively an $\mathcal{M}$-adhesive and an
  $\mathcal{M}'$-adhesive category. Let also $L:\A\rightarrow \C$ be a
  functor that preserves $\mathcal{M}$-pushouts, and
  $R:\B\rightarrow \C$ a functor that preserves pullbacks.Then
  $\comma{L}{R}$ is $\cma{M}{M'}$-adhesive, where
  \[
    \cma{M}{M}':=\{(h,k)\in \mathcal{A}(\comma{L}{R}) \mid h\in
    \mathcal{M}, k\in \mathcal{M}'\}\]
\end{theorem}

@@

Another construction that preserves
$\mathcal{M}$-adhesivity is the formation of the functor category.

\begin{theorem}[\cite{ehrig2006fundamentals,lack2005adhesive}]
  \label{thm:functors}
  Let $\X$ be an $\mathcal{M}$-adhesive category. Then for every small
  category $\Y$, the category $\X^\Y$ of functors $\Y\to \X$ is
  $\mathcal{M}^{\Y}$-adhesive, where
  \[\mathcal{M}^{\Y}:=\{\eta \in \mathcal{A}(\X^\Y) \mid \eta_Y \in
    \mathcal{M} \text{ for every } Y\in \Y\}\]
\end{theorem}

@@

\begin{example}
  \label{ex:adhesive}
  Every topos is adhesive~\cite{lack2006toposes}, thus in particular
  $\cat{Set}$ is adhesive. By the closure properties recalled before
  every presheaf $[\cat{X},\cat{Set}]$ is adhesive and in particular
  the category $\cat{Graph} = [ E \rightrightarrows V, \cat{Set}]$ is
  adhesive where $E \rightrightarrows {V}$ is the two object category
  with two morphisms $s,t \colon{E} \to {V}$. In a similar way,
  various categories of hypergraphs can be shown to be adhesive, while
  the category $\cat{sGraphs}$ of simple graphs, i.e., graphs without
  parallel edges is quasi-adhesive, i.e.,
  $\reg{(\cat{sGraphs})}$-adhesive.\todo{Add hierarchical graphs and term graphs\cite{castelnovo2023thesis,CastelnovoGM22}).}
\end{example}

@@

% \todo[inline]{P: se non impatta sul resto forse sarebbe forse piu' carino definire:\\
%   - for each $G \in \X$, $G : G \Mapsto G$ is a derivation\\  
%   - for each direct derivation $\dder{D}: G \Mapsto H$ and derivation $\der{E} : H \Mapsto F$ we have that $\dder{D}; \der{E} : G \Mapsto F$ is a derivation
% }

@@

\noindent
\parbox{8cm}{
  \begin{definition}[abstraction equivalence]
    Let  $\X$ be an $\mathcal{M}$-adhesive category and 
    $(\X, \R)$ a left-linear DPO rewriting system. 
    An \emph{abstraction equivalence} between two derivations $\der{D}$
    and $\der{D'}$ with the same length and
    $r(\der{D})=r(\der{D}')$, is a family of isomorphisms
    $\{\phi_X\}_{X\in \Deltamin(\der{D})}$ such that the diagram on the right commutes
    for every $i\in [0, \lgh(\der{D})]$. We
    will say that $\der{D}$are $\der{D}'$ are \emph{abstraction
      equivalent} if there exists an abstraction equivalence between
    them.
  \end{definition}
}
\parbox{4cm}{\vspace{-1.35em}$\xymatrix@C=40pt{G'_i&D'_i \ar[r]^{g'_i}
    \ar@{>->}[l]_{f'_i}&G'_{i+1}\\ L_i \ar[u]^{m'_i} \ar[d]_{m_i}& K_i
    \ar[u]^{k'_i} \ar[d]_{k_i} \ar[r]^{r_i} \ar@{>->}[l]_{l_i}
    &R\ar[u]^{h'_i} \ar[d]_{h_i}\\G_i
    \ar@/_.45cm/[uu]_(.35){\phi_{G_i}}|\hole&D_i\ar@{>->}[l]^{f_i}\ar@/_.45cm/[uu]_(.35){\phi_{D_i}}|\hole
    \ar[r]_{g_i}&G_{i+1}\ar@/_.45cm/[uu]_{\phi_{H_i}}}$}

@@

  If $\der{D}=\{\dder{D}_i\}_{i=0}^n$ and
  $\der{D}'=\{\dder{D}'_i\}_{i=0}^n$ with associated sequence of rules
  $r(\der{D})=\{\rho_i\}_{i=0}^n$ and
  $r(\der{D}')=\{\rho'_i\}_{i=0}^n$, a \emph{consistent permutation}
  between $(\der{D}, \alpha, \omega)$ and
  $(\der{D}', \alpha', \omega')$ is a permutation
  $\sigma\colon [0,n]\to [0,n]$ such that, for every $i\in [0,n]$,
  $\rho_i=\rho'_{\sigma(i)}$ and, moreover, there exists a
  \emph{mediating isomorphism}
  $\xi_\sigma\colon \tpro{D} \to \lpro \der{D}' \rpro$ fitting in the
  following diagrams, where $m_i, m'_i, h_i$ and $h'_i$ are,
  respectively, the matches and back-matches of $\dder{D}_i$ and
  $\dder{D}'_i$.

@@

In this section we discuss the notion of independence between
consecutive rewriting steps, which is a fundamental ingredient of the
rewriting theory, which comes into play when viewing a derivation
sequence as a concurrent computation, the idea being that of
abstracting out the order of execution of independent rewritings. We
observe that moving to left-linear rules leads to the failure of
various basic properties of the classical notion of independence and
we single out a framework in which these can be re-established.

@@

It can be shown that switches along the same independence pair are unique up to abstraction
equivalence (see Proposition~\ref{thm:switch_uni}).

As already mentioned, for linear rewriting systems, when viewing
derivations as concurrent computations, it is canonical to identify
derivations which are equal ``up to switching''. The same notion can
be given for left-linear rewriting systems relying on notion of
switch.

@@

In words, we let $\der{D} \equiv^s \der{E}$ when one derivation can be obtained from the other via a sequence of switches.

@@

@@

  A derivation $\der{D}$ consisting of three steps $\dder{D}_0$,
  $\dder{D}_1$, and $\dder{D}_2$, each applying the corresponding rule
  $\rho_i$ ,is depicted in in Fig.~\ref{fi:derD}. The second and third
  steps $\dder{D}_1$ and $\dder{D}_2$ are clearly sequential
  independent, via an independence pairs that map nodes consistently
  with their numbering.\todo{P: having names for the graphs would
    probably help}

@@

  The independence pair can be used to construct a switch $\der{E}$, as
  depicted in Fig.~\ref{fi:derE}.
  \begin{figure}
    \input{Figure/derE.tex}
    \caption{The derivation $\der{E}$.}
    \label{fi:derE}
  \end{figure}
\end{example}

@@

\todo{P: la chiamiamo ``strong indepedence pair''? o ``filler pair''}

\begin{definition}[Good independence pair]
  \label{def:filler}
  Let with $\X$ an $\mathcal{M}$-adhesive category 
  and $(\X, \R)$ a left-linear DPO rewriting system. 
  Let also $(i_0, i_1)$ be an independence pair between two direct
  derivations $\dder{D}_0$, $\dder{D}_1$
  as in the solid part of the diagram below
  \[
    \xymatrix@C=18pt{L_0 \ar[d]_{m_0}&& K_0
      \ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} \ar@{->}@/^.20cm/[ddrr]|(.3)\hole|(.55)\hole_(.7){u_0} & R_0
      \ar@/^.25cm/[drrr]|(.3)\hole^(.2){i_0} \ar[dr]|(.3)\hole_{h_0}
      && L_1 \ar@/_.25cm/[dlll]_(.2){i_1} \ar[dl]|(.3)\hole^{m_1}& K_1
      \ar[d]^{k_1}\ar@{>->}[l]_{l_1} \ar[rr]^{r_1} \ar@{->}@/_.20cm/[ddll]|(.3)\hole|(.55)\hole^(.7){u_1} && R_1 \ar[d]^{h_1} \\
      %
      G_0 &&
      \ar@{>->}[ll]^{f_0} D_0 \ar[rr]_(.3){g_0}&& G_1 &&
      \ar@{>->}[ll]^(.3){f_1} D_1 \ar[rr]_{g_1}&& G_2\\
      %
      & & & & P \ar@/^.3cm/@{>-->}[ull]^{p_0} \ar@/_.3cm/@{-->}[urr]_{p_1}
    }
  \]

  Consider the pullback of $g_0$ and $f_1$, getting $p_i : P \to D_i$
  ($i \in \{0,1\}$) and the mediating arrows $u_0\colon K_0\to P$ and
  $u_1\colon K_1\to P$ into the pullback.
  %given by \Cref{prop:tec} .
  %
  We will say that $(i_0, i_1)$ is a \emph{good independence pair} if
  the resulting squares, depicted  
  below, are pushouts and the pushout of $r_1 : K_1 \to R_1$ and $u_1 : K_1 \to P$ exists.
  \[
    \xymatrix@R=5pt{
      %
      K_0 \ar[r]^{r_0} \ar[d]_{u_0}& R_0 \ar[d]^{i_0} & K_1
      \ar@{>->}[r]^{l_1} \ar[d]_{u_1}& L_1 \ar[d]^{i_1}
      % &K_1 \ar[r]^{r_1} \ar[d]_{u_1}& R_1\ar@{.>}[d]^{j_0}
      \\
      %
      P \ar[r]_{p_1} & D_1 &P \ar[r]_{p_0} & D_0
      % & P \ar@{.>}[r]_{q_1} &
      Q_1}
  \]
\end{definition}

The category $\cat{Graph}$ is tame by the result in ..., while
example~\ref{ex:diff1} shows an adhesive non-tame category.

@@

The correspondence between sequential independence and switchability
is fundamental.  A very large class of DPO rewriting systems - in the
line of \cite{baldan2011adhesivity} - can be identified where all
independence pairs are good for left-linear rewriting systems. The details are in 
\Cref{app:fill}. Here it is important only to observe that such class includes $\cat{Set}$ and it is closed by comma and functor category constructions. As such it includes essentially all categories (e.g., presheaves over set) which are typically considered for modelling purposes.

@@

\begin{restatable}{proposition}{propEqui}
  \label{prop:equi}
  Every linear DPO rewriting system $(\X, \R)$ is tame.
\end{restatable}

[Proof in \Cref{propEqui-proof}]

@@

\begin{definition}[Very tameneness]
	\todo{very tame}
\end{definition}
\todo{SCRIVERE MEGLIO QUESTE RIGHE CON DELLE REFERENZE PERCHÈ IO NON NE SO NIENTE}
\todo{SCEGLIERE ANCHE UN NOME}

The previous definition might sound restrictive. First, it is immediate to see that linear rewriting systems are all very tame (see Proposition~\ref{pr:weak}).
%
Moreover we next introduce a class of rewriting systems, comprising
all the ones used in concrete modelling applications of DPO rewriting
(e.g., in the encoding of process calculi ..., ...) are
actually very tame.

\begin{definition}[Node-merging graphical rewriting systems]
	\todo{sist di riscrittura grafico}
\end{definition}
\todo{esempi (no idea)}
\begin{lemma}
  All node-merging graphical rewriting systems are very tame.
  \todo{tutti very tame}
\end{lemma}

\begin{proof}
  contenuto...
\end{proof}

\begin{example}[E-graphs]
  Graphs with equivalences on nodes. Formally $V \to$
\end{example}

@@

We need two last technical results before proving that  an $\mathcal{M}$-adhesive category of type $\mathbb{B}$ satisfy the $\mathcal{M}$-mixed and $\mathcal{M}$-pushout decomposition properties.

\noindent
\parbox{10.5cm}{\begin{proposition}\label{prop:po2}
		Let $\X$ be an $\mathcal{M}$-adhesive category of type $\mathbb{B}$, and suppose that the square on the right is an $\mathcal{M}$ pullback. Then there exists $E\in \X$, $e\colon E\mto C$ in $\mathcal{M}$, $\phi:E\to B_1$ such that $(C, \{m, e\})$ is a coproduct and $g=f+\phi$. 
\end{proposition}} \parbox{3cm}{\vspace{-.15cm}$\xymatrix{A  \ar[r]^{f} \ar@{>->}[d]_{m}& B_0 \ar@{>->}[d]^{\iota_{B_{0}}}\\ C\ar[r]_-{g} & B_0+B_1}$}
\begin{proof}
		Pulling back $\iota_{B_0}\colon B_0\to B_0\to B_1$ and $\iota_{B_1}\colon B_1\to B_0 + B_1$ we get the diagram below and the thesis follows from \Cref{prop:ext}. The same diagram allows us to deduce $g=f+\phi$.	
	\parbox{5.2cm}{$\xymatrix{A\ar[r]^-{m} \ar@{>->}[d]_{r}& C  \ar@{>->}[d]_{g}& E \ar@{>->}[d]^{\phi}\ar[l]_-{e}\\B_0  \ar[r]_-{\iota_{B_0}}& B_0+B_1 & B_1 \ar[l]^-{\iota_{B_1}} \\ A & B_0 \\C & B_0+E\\ &&B_0+B_1}$}\parbox{8.8cm}{\vspace{-4.2cm}Let us show the second half of the thesis.

	$(\Rightarrow)$ It is immediat to show that the inner square in the second diagram on the left is a pushout. Thus we get the existence of the dotted isomorphism $\varphi\colon B_0+E\to B_0+B_1$.}

@@

\noindent
\parbox{10.7cm}{\begin{corollary}\label{cor:po2}
Let $\X$ be an $\mathcal{M}$-adhesive category of type $\mathbb{B}$, and suppose that the square on the right is an $\mathcal{M}$ pullback. Then there exists $E\in \X$, $e\colon E\mto C$ in $\mathcal{M}$, $\phi:E\to B_1$ such that $(C, \{m, e\})$ is a coproduct and $g=f+\phi$. Moreover, such a square is a pushout if and only if $\phi$ is an isomorphism.
\end{corollary}} \parbox{3cm}{\vspace{-.5cm}$\xymatrix{A  \ar[r]^{f} \ar@{>->}[d]_{m}& B \ar@{>->}[d]^{\iota_{B_{0}}}\\ C\ar[r]_-{g} & B_0+B_1}$}

\iffalse 
\noindent 
\parbox{9.5cm}{\begin{proof}	$(1\Rightarrow 2)$

@@

>>>>

\documentclass[a4paper,UKenglish,cleveref,pdftex, thm-restate,numberwithinsect]{lipics}
\newif\ifreport
\reporttrue
%\reportfalse

% full or conference version
\newif\iffull
\fulltrue
%\fullfalse

% additional text
\iffull
\newcommand{\full}[1]{#1}
%\newcommand{\short}[1]{}
\else 
\newcommand{\full}[1]{}
%\newcommand{\short}[1]{#1}
\fi

@@

\def\R{\mathsf{R}}
\def\B{\textbf {\textup{B}}}
\def\C{\textbf {\textup{C}}}
\def\D{\textbf {\textup{D}}}
\def\X{\textbf {\textup{X}}}
\def\Y{\textbf {\textup{Y}}}
\def\G{\textbf {\textup{G}}}

@@

\title{Left-linear rules in adhesive categories}
\author{Paolo Baldan} 
{Department of Mathematics, University of Padova, Italy}
{baldan@math.unipd.it}{}{}
\author{Davide Castelnovo}
{Department of Mathematics, University of Padova, Italy}
{davide.castelnovo@math.unipd.it}{}{}
\author{Andrea Corradini}
{Department of Computer Science, University of Pisa, Italy}
{andrea.corradini@unipi.it}{}{}

@@

\begin{abstract}
When two sequences of steps executed by a computational device can be considered equivalent?
This is a relevant question if one considers concurrent and distributed systems, since the independence
of two steps means that they could be executed on any order, and potentially in parallel.
%
We investigate the issue in the context of adhesive categories, a very
general instance of the DPO approach to rewriting. More precisely, we
focus on left-linear rules, which not only consume, preserve and generate entities,
as linear rules do, but may ``merge'' parts of the state.
%
An apparently minimal, yet technically hard enhancement, which allows to properly extend the 
formalisms that can be modelled in the adhesive framework, and for which we prove a 
Church-Rosser property.
%
% We refine the established notion of sequential independence between steps, and ...
 \end{abstract}

@@

Here we observe that, in general, both property can fail: sequential
independent steps could be not switchable and different independent
pairs could exist leading to different switches of the same rewriting
steps.  A refined notion of independence can be considered to ensure
that independent rewriting steps can be switched, based in the concept
of a so-called filler. The problem was identified already
in~\cite{baldan2011adhesivity} where a suitable subclass of adhesive
categories is identified where independence as induced by a filler
coincide with sequential independence (as induced by an independence
pair). Here this is extended to $\mathcal{M}$-adhesive categories.

@@

%<<<<<<< HEAD
%\todo{P: qui probabilmente si puo' compattare tenendo solo cio' che si usa}
%\todo{Si usa tutto per la classe B+ [DC]}
%$\mathcal{M}$-adhesivity is well-behaved with respect to  the comma construction \cite{mac2013categories}, as shown by the following theorem.
%\begin{theorem}[\cite{ehrig2006fundamentals,lack2005adhesive}]\label{lem:comma}
  %Let $\A$ and $\B$ be respectively an $\mathcal{M}$-adhesive and an
  %$\mathcal{M}'$-adhesive category. Let also $L:\A\rightarrow \C$ be a
  %functor that preserves $\mathcal{M}$-pushouts, and
  %$R:\B\rightarrow \C$ be a functor which preserves pullbacks.Then
 % $\comma{L}{R}$ is $\cma{M}{M'}$-adhesive, where
  %\[
    %\cma{M}{M}':=\{(h,k)\in \mathcal{A}(\comma{L}{R}) \mid h\in
    %\mathcal{M}, k\in \mathcal{M}'\}\]
%\end{theorem}
$\mathcal{M}$-adhesivity is well-behaved with respect to  the construction of slice and functor caregories, \cite{mac2013categories}, as shown by the following theorems \cite{ehrig2006fundamentals,lack2005adhesive}.
   --ò-à

%\begin{theorem}
%[\cite{ehrig2006fundamentals,lack2005adhesive}]
%\label{lem:comma}
  %Let $\A$ and $\B$ be respectively an $\mathcal{M}$-adhesive and an
 % $\mathcal{M}'$-adhesive category. Let also $L:\A\rightarrow \C$ be a
  %functor that preserves $\mathcal{M}$-pushouts, and
  %$R:\B\rightarrow \C$ a functor that preserves pullbacks.Then
  %$\comma{L}{R}$ is $\cma{M}{M'}$-adhesive, where
  %\[
    %\cma{M}{M}':=\{(h,k)\in \mathcal{A}(\comma{L}{R}) \mid h\in
   % \mathcal{M}, k\in \mathcal{M}'\}\]
%\end{theorem}

@@

\begin{theorem}
  \label{thm:functors}
  Let $\X$ be an $\mathcal{M}$-adhesive category. Then for every small
  category $\Y$, the category $\X^\Y$ of functors $\Y\to \X$ is
  $\mathcal{M}^{\Y}$-adhesive, where
  \[\mathcal{M}^{\Y}:=\{\eta \in \mathcal{A}(\X^\Y) \mid \eta_Y \in
    \mathcal{M} \text{ for every } Y\in \Y\}\]
\end{theorem}

@@

\begin{example}\todo{Topos, ipergrafi e grafi}
\end{example}

\begin{example}
  \todo{grafi semplici}
\end{example}

\begin{example}
  \todo{GRAFI GERARCHICI}
\end{example}

\begin{example}
  \todo{term graph}
\end{example}

@@

\todo[inline]{P: se non impatta sul resto forse sarebbe forse piu' carino definire:\\
  - for each $G \in \X$, $G : G \Mapsto G$ is a derivation\\  
  - for each direct derivation $\dder{D}: G \Mapsto H$ and derivation $\der{E} : H \Mapsto F$ we have that $\dder{D}; \der{E} : G \Mapsto F$ is a derivation
}

@@

\noindent
\parbox{8cm}{
  \begin{definition}
    Let  $\X$ be an $\mathcal{M}$-adhesive category and 
    $(\X, \R)$ a left-linear DPO rewriting system. 
    An \emph{abstraction equivalence} between two derivations $\der{D}$
    and $\der{D'}$ with the same length and
    $r(\der{D})=r(\der{D}')$, is a family of isomorphisms
    $\{\phi_X\}_{X\in \Deltamin(\der{D})}$ such that the diagram on the right commutes
    for every $i\in [0, \lgh(\der{D})]$. We
    will say that $\der{D}$are $\der{D}'$ are \emph{abstraction
      equivalent} if there exists an abstraction equivalence between
    them.
  \end{definition}
}
\parbox{4cm}{\vspace{-1.35em}$\xymatrix@C=40pt{G'_i&D'_i \ar[r]^{g'_i}
    \ar@{>->}[l]_{f'_i}&G'_{i+1}\\ L_i \ar[u]^{m'_i} \ar[d]_{m_i}& K_i
    \ar[u]^{k'_i} \ar[d]_{k_i} \ar[r]^{r_i} \ar@{>->}[l]_{l_i}
    &R\ar[u]^{h'_i} \ar[d]_{h_i}\\G_i
    \ar@/_.45cm/[uu]_(.35){\phi_{G_i}}|\hole&D_i\ar@{>->}[l]^{f_i}\ar@/_.45cm/[uu]_(.35){\phi_{D_i}}|\hole
    \ar[r]_{g_i}&G_{i+1}\ar@/_.45cm/[uu]_{\phi_{H_i}}}$}

@@

  If $\der{D}=\{\dder{D}_i\}_{i=0}^n$ and
  $\der{D}'=\{\dder{D}'_i\}_{i=0}^n$ with assciated sequence of rules
  $r(\der{D})=\{\rho_i\}_{i=0}^n$ and
  $r(\der{D}')=\{\rho'_i\}_{i=0}^n$, a \emph{consistent permutation}
  between $(\der{D}, \alpha, \omega)$ and
  $(\der{D}', \alpha', \omega')$ is a permutation
  $\sigma\colon [0,n]\to [0,n]$ such that, for every $i\in [0,n]$,
  $\rho_i=\rho'_{\sigma(i)}$ and, moreover, there exists a
  \emph{mediating isomorphism}
  $\xi_\sigma\colon \tpro{D} \to \lpro \der{D}' \rpro$ fitting in the
  following diagrams, where $m_i, m'_i, h_i$ and $h'_i$ are,
  respectively, the matches and back-matches of $\dder{D}_i$ and
  $\dder{D}'_i$.

@@

In this section we discuss the notion of independence between
consecutive rewriting steps, which is a fundamental ingredient of the
rewriting theory, which comes into play when viewing a derivation
sequence as a concurrent computation, the idea being that of
abstracting out the order of execution of independent rewritings. We
observe that the shift to left-linear rules leads to the failure of
various basic properties of the classical notion of independence and
we single out a framework in which these can be re-established.

@@

As already mentioned, when viewing derivations as concurrent computations, 
for linear graph rewriting systems it is canonical to identify  derivations which are equal 
``up to switching''. The same notion can be given for left-linear rewriting systems relying on notion of switch.

@@

@@

Switches along the same independence pair are unique up to abstraction
equivalence (see Proposition~\ref{thm:switch_uni}).

@@

  A derivation $\der{D}$ consisting of three steps $\der{D}_0$,
  $\der{D}_1$, and $\der{D}_2$, each applying the corresponding rule
  $\rho_i$ ,is depicted in in Fig.~\ref{fi:derD}. The second and third
  steps $\der{D}_1$ and $\der{D}_2$ are clearly sequential
  independent, via an independence pairs that map nodes consistently
  with their numbering.\todo{P: having names for the graphs would
    probably help}

@@

  The category $\cat{Graph}$ is tame by the result in ...  Thus,
  sequential independence ensure the existence of a switch $\der{E}$,
  depicted in Fig.~\ref{fi:derE}.
  \begin{figure}
    \input{Figure/derE.tex}
    \caption{The derivation $\der{E}$.}
    \label{fi:derE}
  \end{figure}
\end{example}

@@

\begin{definition}[Good independence pair]
  \label{def:filler}
  Let with $\X$ an $\mathcal{M}$-adhesive category 
  and $(\X, \R)$ a left-linear DPO rewriting system. 
  Let also $(i_0, i_1)$ be an independence pair between two direct
  derivations $\dder{D}_0$, $\dder{D}_1$
  as in the solid part of the diagram below
  \[
    \xymatrix@C=15pt{L_0 \ar[d]_{m_0}&& K_0
      \ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} \ar@/^.20cm/[ddrr]|(.3)\hole|(.55)\hole^(.7){u_0} & R_0
      \ar@/^.35cm/[drrr]|(.3)\hole_(.4){i_0} \ar[dr]|(.3)\hole_{h_0}
      && L_1 \ar@/_.35cm/[dlll]^(.4){i_1} \ar[dl]|(.3)\hole^{m_1}& K_1
      \ar[d]^{k_1}\ar@{>->}[l]_{l_1} \ar[rr]^{r_1} \ar@/_.20cm/[ddll]|(.3)\hole|(.55)\hole_(.7){u_1} && R_1 \ar[d]^{h_1} \\
      %
      G_0 &&
      \ar@{>->}[ll]^{f_0} D_0 \ar[rr]_{g_0}&& G_1 &&
      \ar@{>->}[ll]^{f_1} D_1 \ar[rr]_{g_1}&& G_2\\
      %
      & & & & P \ar@/^.3cm/@{>->}[ull]^{p_0} \ar@/_.3cm/[urr]_{p_1}
    }
  \]

  Consider the pullback of $g_0$ and $f_1$, getting $p_i : P \to D_i$
  ($i \in \{0,1\}$) and the arrows $u_0\colon K_0\to P$ and
  $u_1\colon K_1\to P$ given by \Cref{prop:tec} .
  %
  We will say that $(i_0, i_1)$ is a \emph{good independence pair} if
  the first and second square below are pushouts and if the third
  square exists and it is a pushout too
  \[
    \xymatrix{
      %
      K_0 \ar[r]^{r_0} \ar[d]_{u_0}& R_0 \ar[d]^{i_0} & K_1
      \ar@{>->}[r]^{l_1} \ar[d]_{u_1}& L_1 \ar[d]^{i_1}&K_1
      \ar[r]^{r_1} \ar[d]_{u_1}& R_1\ar@{.>}[d]^{j_0} \\
      %
      P \ar[r]_{p_1} & D_1 &P \ar[r]_{p_0} & D_0&P \ar@{.>}[r]_{q_1} &
      Q_1}
  \]
\end{definition}

@@

The correspondence between sequential independence and switchability
is fundamental.  A very large class of DPO rewriting systems - in the
line of \cite{baldan2011adhesivity} - can be identified where all
independence pairs are good for left-linear rewriting systems: ses
\Cref{app:fill}.

@@

\begin{restatable}{proposition}{propEqui}
  \label{prop:equi}
  Every linear DPO rewriting system $(\X, \R)$ is tame.
\end{restatable}
[Proof in \Cref{propEqui-proof}]

@@

\begin{definition}[Very tame rewrityng systems]
A left-linear DPO rewriting system $(\X, \R)$ is \emph{very tame} if it is very tame and, for every derivation $\der{D}:=\{\dder{D}_{i}\}_{i=0}^1$, there is at most one independence paire between $\dder{D}_0$ and $\dder{D}_1$.
\end{definition}

The previous definition might sound restrictive. First, it is immediate to see that linear rewriting systems are all very tame (see Proposition~\ref{pr:weak}).
%
Moreover we next introduce a class of rewriting systems, comprising
all the ones used in concrete modelling applications of DPO rewriting
(e.g., in the encoding of process calculi ..., ...) are \todo{INSERIRE ESEMPI QUI}
actually very tame.

\begin{definition}
Let $\mathcal{G}=(V, E, s, t)$ be a finite directed acyclic graph, we define the \emph{depth} of a vertex $v\in V$ in the following way:
\[\dph(v)=\begin{cases}
0 & \text{there is no $e\in E$ such that $t(e)=v$}\\
m+1 &\text{where } m=\min\{\dph(s(e)) \mid e\in E \text{ such that } t(e)=v \} 
\end{cases}\]
\end{definition}

\begin{remark}
Since $\mathcal{G}$ is acyclic and finite, then $\dph(v)$ is a natural number.
\end{remark}

\begin{definition}[Node-merging graphical rewriting systems]\todo{inserire referenze qui}
Let $\mathcal{G}=(V, E, s, t)$ be a directed acyclic graph, the category  $\gph{G}$ of \emph{$\mathcal{G}$-graphs} is the category $\Set^{\G}$, where $\G$ is the free category over $\mathcal{G}$. A \emph{norde-merging graphical rewriting systems} is a left-linear DPO-rewriting system $(\gph{G}, \R)$ such that, for every rule $(l,r)$, with $l\colon K\to L$ and $r\colon K\to R$  in $\R$, the following hold true:
\begin{enumerate}
	\item for every $v\in V$ and $x\in L(v)$, there exists a vertex $w$ with depth $0$ and an arrow $p\colon w\to v$ (i.e.~a path in $\mathcal{G}$) such that  $x$ belongs to the image of $L(p)\colon L(w)\to L(v)$.
	\item $r\colon K\to R$ is mono on vertices of depth $0$: for every $v\in V$ such that $\dph(v)=0$, the component $r_v:K(v)\to R(v)$ is mono.
\end{enumerate}
\end{definition}

\begin{remark}\label{rem:po}
	Notice that $\gph{G}$ is adhesive and that pushouts are computed componentwise, thus if $g\colon F\to G$ is the pushout of $r\colon K\to R$, then $g_v\colon F(v)\to G(v)$ is mono for every $v$ with depth $0$.
\end{remark}

\begin{example}
\todo{esempi (no idea)}
\end{example}

\begin{restatable}{lemma}{lemVTame}
  All node-merging graphical rewriting systems are very tame.
\end{restatable}

[Proof in \Cref{lemVTame}]

@@

The previous result entails the following one, needed to show that  an $\mathcal{M}$-adhesive category of type $\mathbb{B}$ satisfy the $\mathcal{M}$-mixed and $\mathcal{M}$-pushout decomposition properties.

\noindent
\parbox{10.7cm}{\begin{proposition}\label{prop:po2}
		Let $\X$ be an $\mathcal{M}$-adhesive category of type $\mathbb{B}$, and suppose that the square on the right is an $\mathcal{M}$ pullback. Then there exists $E\in \X$, $e\colon E\mto C$ in $\mathcal{M}$, $\phi:E\to B_1$ such that $(C, \{m, e\})$ is a coproduct and $g=f+\phi$. Moreover, such a square is a pushout if and only if $\phi$ is an isomorphism.
\end{proposition}} \parbox{3cm}{\vspace{-.15cm}$\xymatrix{A  \ar[r]^{f} \ar@{>->}[d]_{m}& B_0 \ar@{>->}[d]^{\iota_{B_{0}}}\\ C\ar[r]_-{g} & B_0+B_1}$}
\iffalse 
\begin{proof}
		Pulling back $\iota_{B_0}\colon B_0\to B_0\to B_1$ and $\iota_{B_1}\colon B_1\to B_0 + B_1$ we get the diagram below and the thesis follows from \Cref{prop:ext}. The same diagram allows us to deduce $g=f+\phi$.	
	\parbox{5.2cm}{$\xymatrix{A\ar[r]^-{m} \ar@{>->}[d]_{r}& C  \ar@{>->}[d]_{g}& E \ar@{>->}[d]^{\phi}\ar[l]_-{e}\\B_0  \ar[r]_-{\iota_{B_0}}& B_0+B_1 & B_1 \ar[l]^-{\iota_{B_1}} \\ A & B_0 \\C & B_0+E\\ &&B_0+B_1}$}\parbox{8.8cm}{\vspace{-4.1cm}Let us show the second half of the thesis.

	$(\Rightarrow)$ It is immediate to show that the inner square in the second diagram on the left is a pushout. Thus we get the existence of the dotted isomorphism $\varphi\colon B_0+E\to B_0+B_1$.}

@@

\noindent
\parbox{10.7cm}{\begin{corollary}\label{cor:po2}
Let $\X$ be an $\mathcal{M}$-adhesive category of type $\mathbb{B}$, and suppose that the square on the right is an $\mathcal{M}$ pullback. Then there exists $E\in \X$, $e\colon E\mto C$ in $\mathcal{M}$, $\phi:E\to B_1$ such that $(C, \{m, e\})$ is a coproduct and $g=f+\phi$. 
\end{corollary}} \parbox{3cm}{\vspace{-.5cm}$\xymatrix{A  \ar[r]^{f} \ar@{>->}[d]_{m}& B \ar@{>->}[d]^{\iota_{B_{0}}}\\ C\ar[r]_-{g} & B_0+B_1}$}

\noindent 
\parbox{9.5cm}{\begin{proof}	$(1\Rightarrow 2)$

@@

\lemVTame*
\label{lemVTame}
